---
title: "dats 6101 project 2 eda"
author: "Xue Ming Wang(G20580112)"
date: "2023-04-16"
output: html_document
---
```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
```

```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```
```{r, results='hide'}
p2 <- read.csv(file = 'data_final.csv')
str(p2)
```
For 15-19 age group
y(Dependent variable):pregnancyrate1519
x(Independent variable): state, abstinence, abortionban_anomaly, abortionban_sex, substance_poli, year, abortionrate1519, birthrate1519,  abortionratio1519, pregnancies1519, abortions1519, births1519, miscarriages1519, population1519, population1544, abortionstotal, birthstotal, pregnancyratetotal, birthratetotal, abortionratetotal, abortionratiototal, miscarriagestotal, pregnanciestotal

For 15 age group
y(Dependent variable): pregnancyratelt15 
x(Independent variable): state, abstinence, abortionban_anomaly, abortionban_sex, substance_poli, year, abortionratelt15, birthratelt15,  abortionratiolt15, pregnancieslt15, abortionslt15, birthslt15, miscarriageslt15, populationlt15, population1544, abortionstotal, birthstotal, pregnancyratetotal, birthratetotal, abortionratetotal, abortionratiototal, miscarriagestotal, pregnanciestotal
```{r, results='hide'}
p2new <- subset(p2, state != "US")
str(p2new)
p2US <- subset(p2, state=="US") 
str(p2US)
```

```{r}
p2new$abstinence=as.factor(p2new$abstinence)

p2new$abortionban_anomaly =as.factor(p2new$abortionban_anomaly)
p2new$abortionban_sex=as.factor(p2new$abortionban_sex)
p2new$substance_poli=as.factor(p2new$substance_poli)
str(p2new)
summary(p2new)
```


# EDA
# US pregnancy rate over years
```{r}
library(ggplot2) 
ggplot(data=p2US, aes(x=year, y=pregnancyrate1519, color = "red")) +
  geom_line()+
  labs(x = "Year", y = "pregnancy rate", title = "US pregnancy rate by Year", caption = "Figure 1. Trend of US pregnancy rate 15-19 over time")
```

# US birth rate over years
```{r}
library(ggplot2) 
ggplot(data=p2US, aes(x=year, y=birthrate1519, color = "red")) +
  geom_line()+
  labs(x = "Year", y = " US birth rate", title = "US birth rate by Year", caption = "Figure 1. Trend of US birth rate 15-19 over time")
```

### Distribution of pregnancyratelt15 
```{r, fig.keep = "none"}
hist(p2$pregnancyratelt15 )
boxplot(p2$pregnancyratelt15 )
qqnorm(p2$pregnancyratelt15 )
qqline(p2$pregnancyratelt15 )
```

### Distribution of pregnancyrate1519 
```{r, fig.keep = "none"}
hist(p2$ pregnancyrate1519 )
boxplot(p2$ pregnancyrate1519 )
qqnorm(p2$ pregnancyrate1519 )
qqline(p2$ pregnancyrate1519 )
```

### Distribution of pregnancyrate40plus 
```{r, fig.keep = "none"}
hist(p2$pregnancyrate40plus )
boxplot(p2$pregnancyrate40plus )
qqnorm(p2$pregnancyrate40plus )
qqline(p2$pregnancyrate40plus )
```

### Distribution of birthrate1519    
```{r, fig.keep = "none"}
hist(p2$birthrate1519)
boxplot(p2$birthrate1519)
qqnorm(p2$birthrate1519)
qqline(p2$birthrate1519)
```
### Distribution of birthratelt15    
```{r, fig.keep = "none"}
hist(p2$birthratelt15)
boxplot(p2$birthratelt15)
qqnorm(p2$birthratelt15)
qqline(p2$birthratelt15)
```

### Distribution of birthrate40plus 
```{r, fig.keep = "none"}
hist(p2$birthrate40plus)
boxplot(p2$birthrate40plus)
qqnorm(p2$birthrate40plus)
qqline(p2$birthrate40plus)
```
# total pregnancies over years
```{r}
library(ggplot2) 
ggplot(data=p2US)+
  geom_point(mapping = aes(x=year, y=pregnancies1519, color=state))+
  labs(title = "Scatterplot of pregnancies in TP", caption = "Figure 7. Scatterplot of pregnancies")
```

# total pregnancy over years in states
```{r}
library(ggplot2) 
ggplot(data=p2new)+
  geom_point(mapping = aes(x=year, y=pregnancies1519, color=state))+
  labs(title = "Scatterplot of pregnancy rate in TP", caption = "Figure 7. Scatterplot of pregnancy rate of each state")
```
# pregnancy rate over years in states
```{r}
library(ggplot2) 
ggplot(data=p2new)+
  geom_point(mapping = aes(x=year, y=pregnancyrate1519, color=state))+
  labs(title = "Scatterplot of pregnancies in TP", caption = "Figure 7. Scatterplot of pregnancies of each state")
```
```{r}
library(maps)
library(ggplot2)

us_map <- map_data("state")

highlight_states <- c("california", "texas", "florida")

ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group, fill = ifelse(region %in% highlight_states, region, "Other")),
               color = "black") +
  theme_void() +
  coord_map() +
  scale_fill_manual(values = c("california" = "red", "texas" = "blue", "florida" = "green", "Other" = "white"), name = "Labels") +
  labs(title = "Outlier States for state birth", caption = "Figure 6. Outlier States for state birth")
```


# pregnancy rate by State Accounting for Abstinence Education
```{r}
library(ggplot2) 
ggplot(data=p2new)+
  geom_point(mapping = aes(x=pregnancyrate1519, y=state, color=abstinence))+
  labs(title = "pregnancy rate by State Accounting for Abstinence Education", caption = "Figure 9. pregnancy rate by State Accounting for Abstinence Education")
```
# cover vs uncovered sex-ed
```{r, results='hide'}
cover <- subset(p2new, abstinence==0) 
str(cover)
notcover <- subset(p2new, abstinence==1) 
str(notcover)
```

# 15-19
```{r}
t_state_birth = t.test(cover$pregnancies1519, notcover$pregnancies1519, conf.level=0.95)
t_state_birth
```

# under 15
```{r}
t_state_birth = t.test(cover$pregnancies1519, notcover$pregnancyratelt15, conf.level=0.95)
t_state_birth
```

# 40+
```{r}
t_state_birth = t.test(cover$pregnancies40plus, notcover$pregnancies40plus, conf.level=0.95)
t_state_birth
```

# subset variables for 15-19 teens
```{r}
p2newdf <- p2new[, c("pregnancyrate1519", "state", "abstinence", "abortionban_anomaly", "abortionban_sex", "substance_poli", "year", "abortionrate1519", "birthrate1519", "abortionratio1519", "pregnancies1519", "abortions1519", "births1519", "miscarriages1519", "population1519", "population1544", "abortionstotal", "birthstotal", "pregnancyratetotal", "birthratetotal", "abortionratetotal", "abortionratiototal", "miscarriagestotal", "pregnanciestotal")]
```

# full model for 1519 teens
```{r}
model1 <- lm(pregnancyrate1519 ~ abstinence+ abortionban_anomaly+ abortionban_sex+ substance_poli+ year+ abortionrate1519+ birthrate1519+  abortionratio1519+ pregnancies1519+ abortions1519+ births1519+ miscarriages1519+ population1519+ population1544+ abortionstotal+ birthstotal+ pregnancyratetotal+ birthratetotal+ abortionratetotal+ abortionratiototal+ miscarriagestotal+ pregnanciestotal + state, data = p2newdf)
summary(model1)
```
full linear model: abortionban_sex1, abortionrate1519, birthrate1519, abortions1519, births1519, population1544, pregnancyratetotal, birthratetotal, stateAR, stateAZ,stateKS, stateKY, stateOK.


# stepwise
```{r}
stepwise_model <- step(model1, direction = "both")
# Print the summary of the stepwise regression model
summary(stepwise_model)
```
stepwise significant varibles: abortionrate1519, birthrate1519, population1544, pregnancyratetotal, birthratetotal, abortionratetotal

# AIC
```{r}
forward_model <- step(model1, direction = "forward")
summary(forward_model)
```
AIC significant varibles: abortionban_sex1, abortionrate1519, birthrate1519, abortions1519, births1519, population1544, pregnancyratetotal, birthratetotal, stateAR, stateAZ,stateKS, stateKY, stateOK.

# BIC
```{r}
backward_model <- step(model1, direction = "backward")
summary(backward_model)
```

BIC significant variables:abortionrate1519, birthrate1519, pregnancyratetotal, birthratetotal abortionratetotal.

# model 2: significant variables 
```{r}
model2 <- lm(pregnancyrate1519 ~ abortionban_sex+ abortionrate1519+ birthrate1519+ abortions1519+ births1519+ population1544+ pregnancyratetotal+ birthratetotal+ abortionratetotal, data = p2newdf)
summary(model2)
```
significant variables: abortionrate1519, birthrate1519, pregnancyratetotal, birthratetotal, abortionratetotal 

# Compare model1 to model2 
```{r}
aov12 <- anova(model1, model2)
aov12
```
model 2 is better 

# model 3: significant variables 
```{r}
model3 <- lm(pregnancyrate1519 ~ abortionrate1519+ birthrate1519+ pregnancyratetotal+ birthratetotal+ abortionratetotal, data = p2newdf)
summary(model3)
```
All varibales significant
```{r}
aov23 <- anova(model2, model3)
aov23
```
model 2 is better 


# correlation matrices
```{r, results='markup'}
library(corrplot)
signp2newdf <- p2new[, c("pregnancyrate1519", "abortionrate1519", "birthrate1519", "pregnancyratetotal", "birthratetotal", "abortionratetotal")]
corPearson = cor(signp2newdf)
corrplot.mixed(corPearson)
# xkabledply(corPearson, title = "Correlation Matrix for bikeshare df, Pearson (Hour==16 subset, all variables)")

corSpearman = cor(signp2newdf, method = "spearman")
corrplot.mixed(corSpearman)
# xkabledply(corSpearman, title = "Correlation Matrix for bikeshare df, Spearman (Hour==16 subset, all variables)")
```
```{r, results="markup"}
corDiff = corPearson - corSpearman
xkabledply(corDiff, title = "Difference between the Pearson and Spearman methods")
corDiffP = 100*corDiff / (corPearson + corSpearman)
xkabledply(corDiffP, title = "Percent Difference between the Pearson and Spearman methods")
```

# model 4 and 5 comparsion 
```{r}
model4 <- lm(pregnancyrate1519 ~ pregnancyratetotal+birthratetotal+abortionratetotal, data = p2newdf)
summary(model4)
model5 <- lm(pregnancyrate1519 ~ pregnancyratetotal*birthratetotal*abortionratetotal, data = p2newdf)
summary(model5)
```

```{r}
aov45 <- anova(model4, model5)
aov45
```
model 5 is better
