---
title: "project1_EDA"
author: "Xue Ming Wang"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
You will also be graded on the technical quality of your analysis as presented in your write-up. You will be assessed for proper and sensical usage of R functions and statistical tools such as:
 Summary of the dataset
 Descriptive statistics
 Graphical representations of the data
 [Proper usage when applicable] Measures of Variance / sd
 [Proper usage when applicable] Normality tests
 [Proper use when applicable] Initial correlation / Chi Square tests / ANOVA analysis / Z-test or Z-interval / T-test or T-interval etc.


```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
```
```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

```{r}
p1 <- read.csv(file = 'Project1_data_clean.csv')
str(p1)
```

## Dataset summary

The dataset for our research includes information on teen birth rates for age group 15-19 for 51 state in the United States (including District of Columbia) from 2005-2015. 

Variable |  Definition  
  :-:    |  :-- 
Year  | The year of the data collection: 2005-2015 
State | 51 state or district of the data
Abstinence.Scales | A score on a scale from 0 to 2, indicating the extent to which abstinence is emphasized in sex education in that state, where 0 means no law mentioned, 1 means covered basic sex education program, and 2 means covered sex education program in strict manner.
Age.Group..Years. | The age group, which is 15-19 years in this dataset
State.Rate | Birth rate for each state for a particular year
State.Births | Number of Birth per State (per 1000)
U.S..Births | Birth rate in U.S (per 1000)
U.S..Birth.Rate | Birth rate for U.S for a particular year 
Unit | The unit of measurement for the state birth and US birth  (per 1000)
Median.Household.Income | The median income of households per state in current dollar
Enrollment | The total number of students enrolled in secondary schools 

## Descriptive statistics

```{r, results='hide'}
p1firsthalf<- data.frame(p1$Year, p1$State, p1$Abstinence.Scales, p1$Age.Group..Years., p1$State.Rate, p1$State.Births)
p2secondhalf <- data.frame( p1$U.S..Births, p1$U.S..Birth.Rate, p1$Unit, p1$Median.Household.Income, p1$Enrollment)
```
The table here represents an overview of all 11 variables.
```{r}
xkablesummary(p1firsthalf, title="Table: Statistics summary")
xkablesummary(p2secondhalf, title="Table: Statistics summary")
```

## Graphical representations of overall trend

### Year vs US Births
```{r}
library(ggplot2) 
ggplot(data=p1, aes(x=Year, y=U.S..Births, color = "red", group=1)) +
  geom_line()+
  geom_point()+
  labs(x = "Year", y = "Number of US Births", title = "US Births by Year", caption = "Figure 1. Trend of US Births over time")
```

The plot (figure 1) shows that the number of US births has declined over time. The number of births peaked in 2007 and reached its lowest point in 2005. The trend presented that US births was generally decreasing from 2005 to 2015.

## Normality tests
We have conducted normality tests on the distributions of the all variables with histogram, boxplot and qq-plot, and identified that two of them are not normally distributed. 

### Distribution of State.Births 
```{r, fig.keep = "none"}
hist(p1$State.Births, main="Histogram of State Birth")
boxplot(p1$State.Births, main="Boxplot of State Birth")
qqnorm(p1$State.Births, main="QQ-plot of State Birth" )
qqline(p1$State.Births )
```
Figure 2. With vs Without Outliers in state birth
```{r}
newStatebirth = outlierKD2(p1, State.Births, rm=TRUE, boxplt=TRUE, histogram=TRUE, qqplt=TRUE)
```

One of the variables is the state birth. In the plots with outleries (Figure 2), it was found to be right-skewed with many outliers. Removing the outliers led to a noticeable improvement in the distribution (Figure 2), although there still appears to be a slight light tail. It may be worth considering further transformations of the variable, like log transformation, to account for the deviation from normality in future research. 

### Distribution of Enrollment 
```{r, fig.keep = "none"}
hist(p1$Enrollment, main="Histogram of Enrollment")
boxplot(p1$Enrollment, main="Histogram of Enrollment")
qqnorm(p1$Enrollment, main="Histogram of Enrollment")
qqline(p1$Enrollment, main="Histogram of Enrollment")
```

Figure 4. With vs Without Outliers in Enrollment
```{r}
newEnrollment = outlierKD2(p1, Enrollment, rm=TRUE, boxplt=TRUE, histogram=TRUE, qqplt=TRUE)
```

Enrollment is another variable that exhibits a non-normal distribution, displaying a right-skewed pattern. However, after the outliers have been addressed, the distribution appears to be more normal than previously observed. It is possible that the presence of outliers in the enrollment data may be attributed to various reasons, such as teens frequently moving in and out of the school. Factors such as school absenteeism, transfer to another school, or re-enrollment could potentially contribute to the occurrence of outliers among teenagers.

### With vs without outliers 
```{r, results='hide'}
# Building new dataframe
Year<-p1$Year
State<-p1$State
Abstinence.Scales<-p1$Abstinence.Scales
State.Rate<-p1$State.Rate
newStatebirth<-newStatebirth$State.Births
U.S..Births<-p1$U.S..Births
U.S..Birth.Rate<- p1$U.S..Birth.Rate
Unit <- p1$Unit 
Median.Household.Income <-p1$ Median.Household.Income
newEnrollment<-newEnrollment$Enrollment

df<- data.frame(Year,State, Abstinence.Scales, State.Rate, newStatebirth, U.S..Births, U.S..Birth.Rate, Unit, Median.Household.Income, newEnrollment)
summary(df)
df<-na.omit(df)
```

```{r}
par(mfrow=c(2,1))
ggplot(p1, aes(x=Year, y=U.S..Birth.Rate))+geom_col(fill="Dark Orange")+
  labs(title = "US Birthrate with Outlier", caption = "Figure 3. US Birthrate with Outlier")
ggplot(df, aes(x=Year, y=newStatebirth))+geom_col(fill="Dark Green")+
labs(title = "US Birthrate without Outlier", caption = "Figure 4. US Birthrate without Outlier")
```
The two plots showed analogous trends, indicating a consistent decrease in the birth rate in the United States from 2007 to 2013, regardless of the presence or absence of outliers.

### Dealing with Outliers

Removing all the outliers can be an option to deal with outliers. On the other hand, it is important to carefully consider the reason for the outliers before removing them, as this could impact the validity of the results.

#### State Birth outliers
```{r, results='hide', fig.keep = 'none'}
newStatebirth= outlierKD2(p1, State.Births, rm=TRUE, boxplt=TRUE, histogram=TRUE, qqplt=TRUE)
outSB = subset(newStatebirth, is.na(State.Births)) 
xkabledplyhead(outSB, title = "Outlier States")
```

```{r}
library(maps)
library(ggplot2)

us_map <- map_data("state")

highlight_states <- c("california", "texas", "florida")

ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group, fill = ifelse(region %in% highlight_states, region, "Other")),
               color = "black") +
  theme_void() +
  coord_map() +
  scale_fill_manual(values = c("california" = "red", "texas" = "blue", "florida" = "green", "Other" = "white"), name = "fill_color") +
  labs(title = "Outlier States for state birth", caption = "Figure 5. Outlier States for state birth")
```

During our examination of state birth outliers, it became evident that the majority of these outliers occur in three particular states: California, Florida, and Texas (Figure 3). In comparison to other states, these three states exhibit substantially higher rates of teen pregnancy. 

If we observe outliers in California, Florida, and Texas with respect to teen pregnancy rates, it is important to note that the high populations of these states could be a contributing factor to the outliers. Therefore, it may not be appropriate to simply delete these outliers without further investigation. Instead, we should consider conducting more in-depth analysis to determine the underlying causes of the high teen pregnancy rates in these states. It is possible that various demographic, socio-economic, or cultural factors could be at play, and understanding these factors could inform targeted interventions to reduce the prevalence of teen pregnancy in these states. For further research on the outliers in California, Florida, and Texas, we would consider incorporating the overall teenage population in each states to see the proportion between teenage population and state birth. This would allow us to better understand the prevalence of teen pregnancy within each state and determine the extent to which it contributes to overall state birth rates.

#### Enrollment outliers
```{r, results='hide', fig.keep = 'none'}
newEnrollment= outlierKD2(p1, Enrollment, rm=TRUE, boxplt=TRUE, histogram=TRUE, qqplt=TRUE)
outE = subset(newEnrollment, is.na(Enrollment)) 
xkabledplyhead(outE, title = "Outlier States")
```

```{r}
library(maps)
library(ggplot2)

us_map <- map_data("state")

highlight_states <- c("california", "texas", "florida")

us_map$fill_color <- ifelse(us_map$region %in% highlight_states, us_map$region, ifelse(us_map$region == "new york", "orange", "white"))

ggplot() +
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group, fill = fill_color),
               color = "black") +
  theme_void() +
  coord_map() +
  scale_fill_manual(values = c("california" = "red", "texas" = "blue", "florida" = "green", "orange" = "orange", "white" = "white")) +
  labs(title = "Outlier States for Enrollment", caption = "Figure 6. Outlier States for Enrollment")
```
During our examination of Enrollment outliers, it became evident that the majority of these outliers occur in four particular states: California, Florida, New York and Texas (Figure 5). In comparison to other states, these three states exhibit substantially higher frequency of enrollment. 

Just like the outliers for state birth, regarding the outliers observed in the enrollment data of California, Florida, New York, and Texas, it may not be appropriate to simply delete them. Instead, it would be better to investigate more on the reasons for these outliers and determine whether they are indicative of any underlying issues related to student mobility or other factors. 
# Unfinished 

## Year vs State Births
```{r}
library(ggplot2) 
ggplot(data=p1)+
  geom_point(mapping = aes(x=Year, y=State.Births, color=State))+
  labs(title = "Outlier States for Enrollment", caption = "Figure 7. Outlier States for Enrollment")
```


## Abstinence Scales

### State.Births vs state
```{r}
library(ggplot2) 
ggplot(data=p1)+
  geom_point(mapping = aes(x=State.Births, y=State, color=Abstinence.Scales))+
  labs(title = "Outlier States for Enrollment", caption = "Figure 5. Outlier States for Enrollment")
```

## difference between Cover vs stress

### split cover and stress into two groups
```{r}
cover <- subset(p1, Abstinence.Scales==1) 
str(cover)
stress <- subset(p1, Abstinence.Scales==2) 
str(stress)
```

### state birth with different Abstinence.Scales 
```{r}
summary(cover$State.Births)
summary(stress$State.Births)
var(cover$State.Births)
var(stress$State.Births)
sd(cover$State.Births)
sd(stress$State.Births)
```

```{r}
Title= c("Min", "1st Qu",  "Median",    "Mean", "3rd Qu",  "Max", "Variance", "SD")
Cover=c(245 ,    868   , 2801 ,   6347 ,   6124,   53417, 116e+08,10769)
Stress=c(530,    3778,    6756,    8515,   10063,   54284, 74614414,8638)
df=data.frame(Title, Cover, Stress)
xkabledply( df, title="Cover vs Stress for State Birth", pos="center", bso="hover" )
```
### state birth rate with different Abstinence.Scales 
```{r, include=F}
summary(cover$State.Rate)
summary(stress$State.Rate)
var(cover$State.Rate)
var(stress$State.Rate)
sd(cover$State.Rate)
sd(stress$State.Rate)
```

```{r}
Title= c("Min", "1st Qu",  "Median",    "Mean", "3rd Qu",  "Max", "Variance", "SD")
Cover=c(10.1,    20.2  ,  28.0,    29.8  ,  38.6,    63.9,142,11.9)
Stress=c(12.1,    25.7,    33.5,    35.4,    43.5,    70.1,152, 12.3)
df=data.frame(Title, Cover, Stress)
xkabledply( df, title="Cover vs Stress for State Birth Rate", pos="center", bso="hover" )
```

### cover and stress in state birth
#### t-test
The null hypothesis here is that, µ1 (cover) = µ0 (stress), the state birth averages are the same. The alternative hypotheses is, µ1 (cover) ≠ µ0 (stress), the state birth averages are different.
```{r}
t_cover_state_birth = t.test(cover$State.Births, stress$State.Births, conf.level=0.95)
t_cover_state_birth
```
The p-value here is 3e-11, which is less than 0.05 and it considered to be statistically significant. Then, the null hypothesis can be rejected. Therefore, µ1 (cover) ≠ µ0 (stress), the state birth averages are different.
 
#### t-interval
```{r}
tint_gpa_cover = t.test(x=cover$State.Births)
tint_gpa_cover
tint_gpa_stress = t.test(x=stress$State.Births)
tint_gpa_stress 
```

### corrlation